---
title: "Ownership crosstabs"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(extrafont)
library(readxl)
library(janitor)
library(lubridate)
library(gmodels)
library(infer)
library(broom)
library(crosstable)

loadfonts()
```

```{r loading jon docs and shapefiles, cache=TRUE, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial"))

regions <- read_csv("Data/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         mif = ifelse(is.na(mif), "TC", mif),
         mif = as.factor(mif),
         mif = fct_relevel(mif, "NW", "NE", "WC", "EC", "SW", "SE", "TC"))

color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

```{r master response, include=FALSE, cache=TRUE}
master.responses <- read_csv("Data/Responses/Master-survey-responses.csv") %>%
  mutate(region = fct_relevel(region, "Northeast", "Northwest", "West Central", "Central", "Southwest", "South Central", "Southeast")) %>%
  rename(public.or.private = 4) %>%
  mutate(public.or.private = str_replace(public.or.private, "Public owned or operated \\(e.g. city, county authority or joint powers board\\)", "Public owned or operated"),
         public.or.private = str_replace(public.or.private, "Private or other non-profit owned or operated", "Private or non-profit"),
         public.or.private = as.factor(public.or.private))
levels(master.responses$public.or.private)
```

This section of analysis examines whether the type of ownership of an agency at all influenced how an agency responded to questions in the survey. The type of ownership question was;

**Which type of ownership best describes your agency?**

Respondents were provided the following options, listed in order of most respondents on top;

* Public (60.8%, 113)
* Private (39.2%, 73)

The metro response was eliminated from analysis.

The survey was broken into 7 different sections. The three sections in bold is what was included  in the analysis due to the MN Department of Health providing incomplete data. You can navigate to the analysis of any of these sections using the table of contents on the left-hand side. The sections are.

1. **Characteristics questions**
2. **Workforce sustainability questions**
3. **Leadership questions**
4. Education questions
5. Medical direction questions
6. Community relation questions
7. Funding questions

The analysis used for the cross tabs is the typical chi-square test with p-value.

<br>

# Summary of results{.tabset}

The tabs below summarize which questions showed a relationship to the type of ownership of an agency and combines the executive summary from the MN Department of Health's analysis with the relationships found in this analysis.

<br>

## Questions with relationships

The following are the questions that showed a statistical significance indicating a relationship between the type of ownership of an agency and how they responded. 

* Characteristics section
  + Is your agency owned, managed or does it receive support from a corporate health system?
  + What type of services do you provide?
  + How are staff compensated, not including the agency director?
  + Are you using Emergency Medical Responders (EMRs) to routinely staff your ambulance(s)?
  + How does your compensation for EMS staff compare with other public safety agencies in your community?

* Workforce sustainability section
  + Is it difficult to recruit new EMS staff?
  + Are any EMS staff cross-trained to work in police, fire, or other public service functions?
  + Which of the following items are obstacles to recruiting EMS staff for your agency? - lack of trained candidates
  + Which of the following items are obstacles to recruiting EMS staff for your agency? - time commitment is too great
  + Which of the following items are obstacles to recruiting EMS staff for your agency? - child care responsibilities
  + Does your EMS agency have a retention strategy?
  + Which of the following items are obstacles to retaining staff for your agency? - excessive time commitment or scheduling conflicts
  + Which of the following items are obstacles to retaining staff for your agency? - limited advancement opportunities
  + Which of the following items are obstacles to retaining staff for your agency? - employer does not allow staff to leave work for a call
  
* Leadership section
  + Has the EMS manager of your agency had formal leadership or management training?

<br>

## Addition to MN Department of Health Summary

The following narrative combines the executive summary from the MN Department of Health's aggregate analysis of the survey with the relationships found in this analysis. Hopefully this will round out the results a bit better and point to more specific trends within the results.

<br>

### Characteristics section{.unnumbered .unlisted .toc-ignore}

**Key Findings**

Most rural EMS agencies provide basic life support services to relatively small populations spread across large geographic areas. Along with low daily call volumes, these realities exacerbate the inability to create sustainable business and staffing models.

* The median call volume is nearly one call per day
* Over half of the agencies cover more than 200 square miles; 37 percent cover more than 300
square miles
* Over half of the agencies serve populations of less than 5,000

However, it is worth noting that there were some differences in responses depending on an agencies type of ownership. 

* *Owned, managed or receive support from a corporate health system: * Although a small percentage of total responses (27%) said they were either owned, managed or receive support from a corporate healthy system, agencies that were privately owned or are a non-profit confirmed this relationship at a significantly higher than publicly owned agencies (58% vs. 9%).
* *Types of services provided: * although a very large percentage of agencies responded that they provide basic life support (88.2% of total responses), publicly owned agencies provided this service at a significantly higher rate than privately owned or non-profit agencies (97% vs. 78%). On the flip side, privately owned or non-profit agencies responded that they provide advanced life support at a higher rate than publicly owned agencies (51% vs. 12%).
* *Staff compensation: * although a majority of agencies responded that they pay their staff using a volunteer / on call compensation structure (57% of total responses), publicly owned agencies were much more likely to pay this way than privately owned or non-profit agencies (67% vs. 39%). On the flip side, agencies that were privately owned or a non-profit confirmed that they pay their staff at a significantly higher rate than publicly owned agencies (42% vs. 6%).
* *Using EMRs to staff ambulance: * Although a majority of agencies (59%) responded that they do NOT use EMRs to staff their ambulance(s), privately owned or non-profit agencies responded this way at a significantly higher rate than publicly owned agencies (78% of responses vs. 46%). 
* *Compensation comparison to other public safety agencies: * Although a very small percentage (13%) of agencies said they paid their staff better compared to other public safety agencies, privately owned or non-profit agencies responded this way at a significantly higher rate than publicly owned agencies (22% vs. 8%).

<br>

### Workforce sustainability section{.unnumbered .unlisted .toc-ignore}

**Key Findings**

Staffing shortages are a major challenge for rural EMS agencies. Eighty (80) percent utilize some version of volunteers to staff their operations. Roster sizes are decreasing, many staff are trained for multiple public functions and many call schedules are not covered 24 hours prior to the shifts.

* The active staff roster decreased for half of the agencies from the previous year
* Weekday daytime hours are the most difficult shifts to cover. However, all weekend and holiday
shifts are also difficult to cover
* About 60 percent of agencies have inadequate staff to cover their call schedule without undue
burden to the agency.
* Fifty-nine (59) percent of agencies do not have their shifts 100 percent covered at least 24 hours in
advance

Recruitment and retention of ambulance personnel continues to be a significant challenge even though:
* Emergency Medical Responders (EMRs) are now widely used to actively staff rural EMS agencies
* Most agencies cover the full cost of the continuing education courses for their staff
* Most eligible recipients take advantage of state EMS training reimbursement

However, it is worth noting that there were some differences in responses depending on an agencies type of ownership. 

* *Difficult to recruit new staff: * Although a majority of agencies (63%) said that staff recruitment was difficult, publicly owned agencies responded this way at a significantly higher rate than privately owned or non-profit agencies (70% vs. 51%). 
* *Cross-training of EMS staff: * Although a large majority of agencies (74%) said that their EMS staff were cross-trained in other publicly safety positions, publicly owned agencies responded this way at a significantly higher rate than privately owned or non-profit agencies (85% vs. 60%). 
* *Obstacles to recruiting - lack of trained candidates: * Although half of all agencies (50%) responded that a lack of trained candidates was a barrier to staff recruitment, it was a more significant issue for privately owned or non-profit agencies (66% of these agencies vs. 41% of publicly owned agencies). 
* *Obstacles to recruiting - time commitment is too great: * Although a majority of all agencies (67%) responded that the time commitment being too great as a barrier to staff recruitment, it was a more wide-spread issue among publicly owned agencies (81% of these agencies vs. 47% of privately owned or non-profit agencies). 
* *Obstacles to recruiting - child care responsibilities: * Although only 38% of all agencies said that child care responsibilities were a barrier to staff recruitment, it was a more widespread issue for publicly owned agencies (48% of these agencies vs. 28% of publicly owned or non-profit agencies).
* *Agency have a retention strategy: * Half of all agencies responded that they have a retention strategy, however privately owned or non-profit agencies had one at a much higher rate than publicly owned agencies (68% vs. 41% respectively).
* *Obstacles to retention - excessive time commitment: * Although a large majority of all agencies said that the excessive time commitment or scheduling conflicts were a barrier to staff retention, publicly owned agencies confirmed this barrier at a significantly higher rate than privately owned or non-profit agencies (82% vs. 57%)
* *Obstacles to retention - limited advancement opportunities: * Although only 23% of all agencies said that limited advancement opportunities were a barrier to staff retention, privately owned or non-profit agencies confirmed this barrier at a significantly higher rate than publicly owned agencies (42% vs. 12%).
* *Obstacles to retention - employer not allowing staff to leave: * Although only 37% of all agencies said that employers not allowing staff to leave work for a call was a barrier to staff retention, publicly owned agencies confirmed this barrier at a significantly higher rate than privately owned or non-profit agencies (46% vs. 26%).

<br>

### Leadership section{.unnumbered .unlisted .toc-ignore}

**Key Findings**

The largest problems that rural EMS agency managers face today, as in the past, are recruitment of staff and filling call schedules. These two management tasks seem to form a Maslow’s Hierarchy of basic management needs (see Appendix C, Question 24). It may be that efforts to address other management issues will be fruitless and/or unwelcomed until these basic needs are addressed.

* Sixty-nine (69) percent of rural EMS managers report difficulty recruiting staff despite the fact that 57.5 percent have a recruitment plan
* Over half of rural EMS managers report difficulty staffing their schedule

High leadership turnover was defined in this survey as an EMS agency having had three or more managers in the past five years. Using this definition, there is remarkably low turnover of rural EMS managers. A high majority of EMS managers have an active role in developing their budgets, but generally are not directly involved in billing.

* Eighty-six (86) percent of services did not have high turnover of their managers
* Eighty-one (81) percent of rural EMS managers have a role in developing their annual budgets
* Eighty-one (81) percent use other resources to bill, with billing agencies as the most common at
fifty-two (52) percent

However, it is worth noting that there was one difference in responses depending on the type of ownership. 

* *Management received training: * Although a majority of agencies (54%) responded that their management has received formal leadership or management training, privately owned or non-profit agencies responded this way at a significantly higher rate than publicly owned agencies (67% vs. 46%).



<br>

# Characteristics section analysis{.tabset}

The questions in the demographic section of the survey are provided in tabs below along with the chi-square and p-value for the relationship between region and each question.

The questions where type of ownership had a relationship in how an agency responded are summarized below follow by the analysis and contingency tables.

**Summary:** 

The big take-away is that the type of ownership influenced nearly all of these questions in a very consistent way. Private/non-profit ownership typically meant that they were more likely to respond to the following questions at a higher rate;
* that they receive corporate health support, 
* they offer advanced life support, 
* pay their staff and were compensated better or the same as other public safety positions, and,
* didn't use EMS to staff their ambulances.

The opposite was true for publicly owned agencies. This type of ownership meant that they responded in the following way at a significantly higher rate than expected;
* did not receive corporate health support,
* offered basic life support over advanced life support,
* did not pay staff and were compensated worse than other public safety positions, and,
* used EMS to staff their ambulances.

<br>

## Corporate health support

**Is your agency owned, managed or does it receive support from a corporate health system?**

Respondents were provided two choices which are provided below and in order of the top one receiving the most responses;

* No (72%)
* Yes (27%)


Probably not surprising that private/non-profit agencies responded that they received corporate health support at a significantly higher rate compared to publicly owned agencies - 58% vs. 1% respectively.

<br>

```{r ownership and corporate health support}
ownership.corp.health.support <- master.responses %>%
  select(4,9) %>%
  rename(corp.health.support = 2)

CrossTable(ownership.corp.health.support$public.or.private, ownership.corp.health.support$corp.health.support, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
  

```

## Services provided{.tabset}

**What types of services do you provide?**

This question provided 4 choices which are provided below in order of the top one receiving the most responses;

* Basic Life Support (88%)
* Advacned life support (27%)
* Part-time advanced life support (17%)
* Specialized advanced life support (9%)

<br>

```{r prep ownership and services provided}
ownership.services <- master.responses %>%
  select(4, 10:13) %>%
  gather(key = "services", value = "provided", 2:5) %>%
  mutate(provided = as_factor(provided),
         services = str_replace(services, "ServicesProvided \\(Q11_1\\)", "Basic life support"),
         services = str_replace(services, "ServicesProvided \\(Q11_2\\)", "Advanced life support"),
         services = str_replace(services, "ServicesProvided \\(Q11_3\\)", "Part-time advanced life support"),
         services = str_replace(services, "ServicesProvided \\(Q11_4\\)", "Specialized life support"),
         services = as.factor(services),
         provided = ifelse(provided == "-", "No", "Yes"))

ownership.als <- ownership.services %>%
  filter(services == "Advanced life support")

ownership.bls <- ownership.services %>%
  filter(services == "Basic life support")

ownership.parttime.als <- ownership.services %>%
  filter(services == "Part-time advanced life support")

ownership.sls <- ownership.services %>%
  filter(services == "Specialized life support")

```

### Advanced life support

There is a relationship in whether an agency offers advanced life support and the type of ownership.


Agencies that were privately or other non-profit owned or operated offered advanced life support at a higher rate than expected - 38 (51%) responses compared to the expected response of 21 (27%). The opposite is true for publicly owned - they had a significantly lower number of responses - 14 (12%) responded vs the expected 31 (27%).

<br>

```{r advanced life support contingency table}

CrossTable(ownership.als$public.or.private, ownership.als$provided, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Basic life support

There is a relationship in whether an agency offers basic life support and their type of ownership, although not surprising considering the previous question. 

The publicly owned agencies responded that they provided basic life support services at a much higher rate than expected - 110 (97%) vs. the expected 100 (88%). The opposite is true for private or non-profit owned agencies - 56 (76%) responses vs. the expected 66 (88%).

<br>

```{r basic life support contingency table}

CrossTable(ownership.bls$public.or.private, ownership.bls$provided, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Part-time Advanced life support

There is no relation in whether an agency offers part-time life support and their type of ownership

<br>

```{r part-time advanced life support contingency table}

CrossTable(ownership.parttime.als$public.or.private, ownership.parttime.als$provided, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Specialized life support

There is no relationship in whether an agency offers specialized life support and their type of ownership.

<br>

```{r specialized life support contingency table}

CrossTable(ownership.sls$public.or.private, ownership.sls$provided, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

## Staff compensation

**How are staff compensated, not including the agency director?**

Respondents were provided 4 choices which are provided below and ranked with top choice receiving the most responses;

* Volunteer, on call compensation (57%)
* Paid, hourly wage or salary (20%)
* Combination, volunteer and paid (17%)
* Volunteer, no compensation (5%)

There is a relationship in the type of compensation provided and type of ownership.

EMS agencies that were private or non-profit responded that they paid their staff at a higher rate than expected - 31 responses (42%) vs. the 15 (20%) expected. The opposite is true for publicly owned agencies. They responded at a lower rate with 7 (1%) responding that they paid their staff vs the 23 (20%) expected.


<br>

```{r ownership and compensation}
own.comp <- master.responses %>%
  select(4,14) %>%
  rename(compensation = 2) %>%
  mutate(compensation = str_replace(compensation, "Combination of volunteer and paid staff", "Paid and vol"),
         compensation = str_replace(compensation, "Paid; hourly wage or salary", "Paid"),
         compensation = str_replace(compensation, "Volunteer \\(receive no compensation at all\\)", "Vol"),
         compensation = str_replace(compensation, "Volunteer/Paid; on-call compensation \\(on-call compensation\\); meets statutory definition of volunteer https://www.revisor.mn.gov/statutes/\\?id=144E.001#stat.144E.001.15", "On-call"),
         compensation = as.factor(compensation))

CrossTable(own.comp$public.or.private, own.comp$compensation, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE, max.width = 2)
  

```

<br>

## Using EMS to staff ambulances

**Are you using Emergency Medical Responders (EMRs) to routinely staff your ambulance(s)?**

Respondents were provided two choices which are provided below in order of the selection with the most responses on top; note: it was incorrect in their report. The answers were flipped.

* Yes (41%)
* No (59%)

There is a significant relationship in whether an agency uses EMS to staff their ambulance and type of ownership.

Agencies that are private or a non-profit responded that they were utilizing EMS to staff their ambulances at a significantly lower rate - 16 responses 21% vs the 30 (41%) expected. The opposite is true for publicly owned agencies. They were using EMS to staff their ambulances at a much higher response rate - 61 responses (54%) vs. the 47 expected (41%).

<br>

```{r ownership and ems for staffing}
ownership.ems <- master.responses %>%
  select(4,21) %>% 
  rename(ems.use = 2) 
  
CrossTable(ownership.ems$public.or.private, ownership.ems$ems.use, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
  

```

<br>

## Compensation comparison

**How does you compensation for EMS staff compare with other public safety agencies in your community?**

Respondents were provided 4 choices, which are listed below in order of the top choice on top;

* Worse (33%)
* Same (29%)
* Unknown (24%)
* Better (13%)

There is a relationship in compensation comparison and type of ownership.

Agencies that were private or non-profit responded that they paid their staff better or the same in comparison of other safety agencies at a higher rate than expected - Better: 16 responses (21%) vs 10 expected (13%); Same: 26 responses (35%) vs. 22 expected (29%). 

The opposite is true for publicly owned agencies. They responded that they paid their staff better or the same as other public safety agencies at a lower rate than expected - Better: 9 responses (1%) vs 15 expected (13%), Same: 29 responses (26%) vs. 33 expected (29%).

<br>

```{r ownership and compensation comparison for staffing}
ownership.comp.comp <- master.responses %>%
  select(4,22) %>%
  rename(comp.comp = 2) 

CrossTable(ownership.comp.comp$public.or.private, ownership.comp.comp$comp.comp, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
  

```

<br>

# Workforce sustainability section analysis{.tabset}

The questions in the workforce sustainability section of the survey are provided in tabs below along with the chi-square and p-value for the relationship between region and each question.

The questions where region had a relationship in how an agency responded are summarized below.

**Summary:**

*Difficult to staff recruitment:* 
Agencies that were private or non-profit responded that recruitment was difficult at a lower rate compared to what was expected. The opposite was true for publicly owned agencies - they responded that recruitment was difficult at a much higher rate.

*Cross-training of staff*
Publicly owned agencies responded that there staff was cross-trained in other public safety agencies at a higher rate than expected. For private/non-profit agencies, the opposite was true.

*Obstacles to recruiting:* 
The main obstacles where there seemed to be an influence in responses due to ownership type were;

* Lack of trained candidates: private/non-profit agencies responded that this was a barrier to recruitment at a higher rate than expected.
* Too much time commitment: Publicly owned agencies responded that this was a barrier to recruitment at a higher rate than expected.
* Child-care obligations: Publicly owned agencies responded that this was a barrier to recruitment at a higher rate than expected.

*Retention strategy*
Private/non-profit agencies responded that they had a retention strategy at a higher rate than expected. The opposite is true for publicly owned agencies.


*Obstacles to retention:* 
The main obstacles where there seemed to be an influence in responses due to ownership type were;

* Time commitment: Publicly owned agencies responded that this was a barrier to retention at a higher rate than expected.
* Limited advancement: Private/non-profit agencies responded that this was a barrier to retention at a higher rate than expected.
* Employer won't allow staff to leave work: Publicly owned agencies responded that this was a bartrier to retention at a higher rate than expected.

<br>

## Cover call schedule

**Do you have enough people to cover your call schedule without undue burden or excessive time commitments?**

Respondents were provided with two choices which are provided below in order of top choice listed first;

* No (60%)
* Yes (40%)

There is a relationship in whether an agency has their call schedule covered and type of ownership.

Private/non-profit agencies responded that they were able to cover the call schedule without undue burden or excessive time commitments at a higher rate than expected - 38 responded (51%) vs. the 29 expected (40%).

Publicly owned agencies responded that they could cover the call schedule at a lower rate than expected - 36 responded yes (32%) vs. the 45 expected (40%).

<br>

```{r ownership and covering all schedule}
owner.schedule <- master.responses %>%
  select(4,23) %>%
  rename(cover.sched = 2) 

CrossTable(owner.schedule$public.or.private, owner.schedule$cover.sched, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
  

```

<br>

## Roster changes

**Over the past year, has the number of actie staff on your roster changed?**

Respondents were provided three choices which are provided below in order of top choice listed first;

* Decreased (50%)
* Same (29%)
* Increased (21%)

There is no relationship in changes to roster at an agency and their type of ownership

<br>

```{r ownership and roster changes}

owner.rost <- master.responses %>%
  select(4,24) %>%
  rename(rost.changes = 2) %>%
  mutate(rost.changes = str_replace(rost.changes, "Size of the active staff has not changed", "No change"),
         rost.changes = as.factor(rost.changes))

CrossTable(owner.rost$public.or.private, owner.rost$rost.changes, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
  

```

<br>

## Recruitment plan

**Does you agency have a recruitment plan?**

Respondents were provided two choices which are provided below in order of top choice listed first;

* Yes (58%)
* No (41%)

There is no relationship in whether an agency has a recruitment plan and their type of ownership.

<br>

```{r owernship and recruitment plan}

owner.rec.plan <- master.responses %>%
  select(4,25) %>%
  rename(rec.plan = 2) 

CrossTable(owner.rec.plan$public.or.private, owner.rec.plan$rec.plan, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
  

```

<br>

## Difficult to recruit?

**Is it difficult to recruit new EMS staff?**

Respondents were provided three choices which are list below with the top choice first;

* Yes (63%)
* No (4%)
* Sometimes (33%)

There is a relationship in whether an agency feels it's difficult to recruit new staff and their region.

Agencies that were privately owned or a non-profit responded that recruitment was difficult at a lower rate than expected - 38 responses (51%) vs. 46 expected (63%). The opposite was true for publicly owned - 79 responses (71%) that it was difficult to recruit vs. the 71 expected (63%).

<br>

```{r ownership and difficult to recruit}

owner.dif.rec <- master.responses %>%
  select(4,26) %>%
  rename(dif.rec = 2) 

CrossTable(owner.dif.rec$public.or.private, owner.dif.rec$dif.rec, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
  

```

<br>

## Cross-training of staff

**Are any EMS staff cross-trained to work in police, fire, or other public service functions?**

Respondents were provided two choices which are listed below with the top choice first;

* Yes (74%)
* No (25%)

There is a relationship in whether an agency has staff cross-trained in other public emergency agencies and type of ownership.

Privately owned or non-profit agencies responded that their staff was cross-trained at a significantly lower rate than expected - 43 responses (59%) vs. 55 expected (74%). The opposite was true for publicly owned agencies - 95 responded (85%) that were cross-trained vs. the 84 expected (74%).

<br>

```{r ownership and cross-training}

owner.cross.train <- master.responses %>%
  select(4,27) %>%
  rename(cross.train = 2) 

CrossTable(owner.cross.train$public.or.private, owner.cross.train$cross.train, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
  

```

<br>

## Obstacles to recruiting{.tabset .tabset-pills}

**Which of the following items are obstacles to recruiting EMS staff for your agency?**

Respondents were provided 10 choices of which they coud select all that applied. The choices are provided below and in order with top choice listed first;

* Time commitment is too great (67%)
* Lack of availability (62%)
* Training requirements are too time consuming (58%)
* Lack of trained candidates (50%)
* inadequate pay or benefits (40%)
* Child care responsibilities (38%)
* Training requirements are too far away (23%)
* Too few runs or transports (14%)
* Too many runs or transports (9%)
* Other (5%)

There was a relationship between type of ownership and the following barriers to recruitment;

* Lack of trained candidates
* Too much time commitment
* Child care obligations

<br>

```{r prep ownership and obstables to recruiting, include=FALSE, cache=TRUE}

owner.obs.rec <- master.responses %>%
  select(4,28:38) %>%
  rename(lack.interest = 2,
         lack.availability = 3,
         lack.trained.cand = 4,
         training.too.time = 5,
         training.too.exp = 6,
         training.too.dist = 7,
         inadequate.pay = 8,
         too.few.runs = 9,
         too.many.runs = 10,
         too.time.commit = 11,
         child.care = 12) %>%
  gather(key = "obstacle", value = "response", 2:12) %>%
  mutate(response = ifelse(response == "-", "No", "Yes"))
```

### Lack of interest

There is no relationship between type of ownership and whether lack of interest is considered an obstacle to recruitment.

<br>

```{r ownership and obstables to recruiting - lack of interest}
owner.obs.rec.int <- owner.obs.rec %>%
  filter(obstacle == "lack.interest")

CrossTable(owner.obs.rec.int$public.or.private, owner.obs.rec.int$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Lack of availability

There is no relationship between type of ownership and whether lack of availability was considered an obstacle to recruitment.

<br>

```{r ownership and obstables to recruiting - lack of availability}
owner.obs.rec.availability <- owner.obs.rec %>%
  filter(obstacle == "lack.availability")

CrossTable(owner.obs.rec.availability$public.or.private, owner.obs.rec.availability$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Lack of trained candidates

There is a relationship between type of agency and whether there is a lack of trained candidates acting as an obstacle to recruitment.

Private or non-profit agencies responded that a lack of trained candidates was an obstacle to recruitment at a higher rate than expected - 49 responses (66%) vs 38 expected (50%). The opposite is true for publicly owned - 46 responded (41%) that a lack of trained candidates was an obstacle vs. the 57 expected (50%).

<br>

```{r ownership and obstables to recruiting - lack of candidates}
owner.obs.rec.trained.cand <- owner.obs.rec %>%
  filter(obstacle == "lack.trained.cand")

CrossTable(owner.obs.rec.trained.cand$public.or.private, owner.obs.rec.trained.cand$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Training takes too much time

There is no relationship between type of ownership and whether training takes too much time and acts as an obstacle to recruitment.


<br>

```{r ownership and obstables to recruiting - training too much time}
owner.obs.rec.training.time <- owner.obs.rec %>%
  filter(obstacle == "training.too.time")

CrossTable(owner.obs.rec.training.time$public.or.private, owner.obs.rec.training.time$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Training too expensive

There is no relationship between type of ownership and whether trainings being too expensive was an obstacle to recruitment.

<br>

```{r ownership and obstables to recruiting - training too expensive}
owner.obs.rec.training.exp <- owner.obs.rec %>%
  filter(obstacle == "training.too.exp")

CrossTable(owner.obs.rec.training.exp$public.or.private, owner.obs.rec.training.exp$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Training locations too far

There is no relationship between type of ownership and whether trainings were too far to travel were obstacles to recruitment.

<br>

```{r ownership and obstables to recruiting - training too far}
owner.obs.rec.trainings.too.far <- owner.obs.rec %>%
  filter(obstacle == "training.too.dist")

CrossTable(owner.obs.rec.trainings.too.far$public.or.private, owner.obs.rec.trainings.too.far$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Inadequate pay

There is no relationship between type of agency and whether inadequate pay was an obstacle to recruitment.

<br>

```{r ownership and obstables to recruiting - inadequate pay}
owner.obs.rec.inadequate.pay <- owner.obs.rec %>%
  filter(obstacle == "inadequate.pay")

CrossTable(owner.obs.rec.inadequate.pay$public.or.private, owner.obs.rec.inadequate.pay$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Too few runs

There is no relationship between type of agency and whether too few runs is an obstacle to recruitment.

<br>

```{r ownership and obstables to recruiting - too few runs}
owner.obs.rec.too.few.runs <- owner.obs.rec %>%
  filter(obstacle == "too.few.runs")

CrossTable(owner.obs.rec.too.few.runs$public.or.private, owner.obs.rec.too.few.runs$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Too many runs

There is no relationship between type of agency and whether too many runs was an obstacle in recruitment.

<br>

```{r ownership and obstables to recruiting - too many runs}
owner.obs.rec.too.many.runs <- owner.obs.rec %>%
  filter(obstacle == "too.many.runs")

CrossTable(owner.obs.rec.too.many.runs$public.or.private, owner.obs.rec.too.many.runs$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Too much time commitment

There is a relationship between type of ownership and whether too much of a time commitment was an obstacle to recruitment.

Private or non-profit agencies responded that the time commitment wasn't an obstacle to recruitment at a much higher rate than expected - 39 responses (53%) vs. 24 expected (33%). The opposite is true for publicly owned agencies - 91 responded (81%) that the time commitment was an obstacle to recruitment vs 76 expected (67%).

<br>

```{r ownership and obstables to recruiting - too much time commitment}
owner.obs.rec.too.much.time <- owner.obs.rec %>%
  filter(obstacle == "too.time.commit")

CrossTable(owner.obs.rec.too.much.time$public.or.private, owner.obs.rec.too.much.time$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

### Child care obligations

There is a relationship between type of ownership and whether child care obligations were an obstacle to recruitment.

Private or non-profit agencies responded that childcare obligations were not an obstacle to recruitment at a higher rate than expected - 53 responses (72%) vs. 44 expected (62%). The opposite is true for publicly owned agencies. These agencies responded that child care obligations were an obstacle to recruitment at a higher rate than expected - 54 responses (48%) vs. 45 expected (38%).

<br>

```{r ownership and obstables to recruiting - child care}
owner.obs.rec.child.care <- owner.obs.rec %>%
  filter(obstacle == "child.care")

CrossTable(owner.obs.rec.child.care$public.or.private, owner.obs.rec.child.care$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>


## Retention strategy

**Does your EMS agency have a retention strategy?**

Respondents were given two options which are provided below in order of top choice first;

* Yes (50%)
* No (49%)

There is a relationship between type of ownership and whether an agency has a retention strategy. 

Private or non-profit agencies responded that they have a retention strategy at a higher rate than expected - 50 responses (68%) vs. 38 expected (50%). The opposite is true for publicly owned agencies. They responded that they did not have a retention strategy at a higher rate than expected - 66 responses (60%) vs. 54 expected (49%).

<br>

```{r ownership and retention strategy}
own.retention.strat <- master.responses %>%
  select(4,41) %>%
  rename(retention.strat = 2)

CrossTable(own.retention.strat$public.or.private, own.retention.strat$retention.strat, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

## Trouble retaining staff

**Does your agency have difficulty retaining existing staff?**

Respondents were provided three choices which are below in order of top choice listed first;

* Sometimes (84%)
* Always (10%)
* Never (5%)

There is no relationship between the type of ownership and their response to if they were having difficulty retaining staff. 

<br>

```{r ownership and difficulty retaining staff}
own.retaining.staff <- master.responses %>%
  select(4, 42) %>%
  rename(retaining.staff = 2)

CrossTable(own.retaining.staff$public.or.private, own.retaining.staff$retaining.staff, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

## Obstacles to retaining{.tabset .tabset-pills}

**Which of the following items are obstacles to retaining staff for your agency?**

This question provided a number of choices for an agency to select and were allowed to check all that apply. The choices are provided below in order of the choice receiving the most responses at the top and working our way down. The value in parenthesis is the percentage of responses it received.

* Excessive time commitment or scheduling conflicts (72%)
* Age: older crew members retiring (60%)
* Inadequate pay or benefits (39%)
* Employer does not allow staff to leave work for a call (37%)
* Certification is too difficult to maintain (26%)
* Limited advancement opportunities (23%)
* Lack of recognition (14%)
* Too many runs or transports (11%)
* Community is too isolated (10%)
* Other (10%)
* Too few runs or transports (10%)
* Dissatisfaction with EMS administrator/owner (5%)
* Dissatisfaction with job duties (3%)
* Dissatisfaction with EMS manager (1%)


The following barriers had a relationship to the type of ownership;

* Time commitment
* Limited advancement
* Employer won't allow staff to leave work

<br>

```{r prep ownershop and obstacles to retention, include=FALSE, cache=TRUE}
own.retention.obstacles <- master.responses %>%
  select(4,43:55)  %>%
  rename(retirement = 2,
         time.committment = 3,
         certification.maintain = 4,
         inadequate.pay = 5,
         limited.advancement = 6,
         lack.recognition = 7,
         community.too.isolated = 8,
         employer.flexibility = 9,
         too.many.runs = 10,
         too.few.runs = 11,
         dissatisfaction.owner = 12,
         dissatisfaction.manager = 13,
         dissatisfaction.duties = 14) %>%
  gather(key = "obstacles", value = "response", 2:14) %>% 
  mutate(response = ifelse(response == "-", "No", "Yes"))
```

<br>

### Retirement

There is no relationship between type of ownership and an agency's response to retirement being an obstacle to retention. 

<br>

```{r ownership and retention - retirement}
own.retention.retirement <- own.retention.obstacles %>%
  filter(obstacles == "retirement")

CrossTable(own.retention.retirement$public.or.private, own.retention.retirement$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Time committment

There is a relationship between type of ownership and an agency's response to whether excessive time commitment was an obstacle to retainment.

Private or non-profit agencies responded that time commitment was not a barrier to retention at a higher rate than expected - 32 responses (43%) vs. 21 expected (28%). The opposite was true for publicly owned agencies. 93 (82%) publicly owned agencies responded that time commitment was a barrier to retainment vs. the 82 expected (72%).

<br>

```{r ownership and retention - time committment}
own.retention.time.commit <- own.retention.obstacles %>%
  filter(obstacles == "time.committment")

CrossTable(own.retention.time.commit$public.or.private, own.retention.time.commit$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Difficult to maintain certification

There is no relationship between type of ownership and whether an agency indicated that difficulty to maintain certification was an obstacle to retainment.

<br>

```{r ownership and retention - maintain certification}
own.retention.maintain.certification <- own.retention.obstacles %>%
  filter(obstacles == "certification.maintain")

CrossTable(own.retention.maintain.certification$public.or.private, own.retention.maintain.certification$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Inadequate pay

There is no relationship between type of ownership and whether an agency indicated that inadequate pay was an obstacle to retainment.

<br>

```{r ownership and retention - inadequate pay}
own.retention.inadequate.pay <- own.retention.obstacles %>%
  filter(obstacles == "inadequate.pay")

CrossTable(own.retention.inadequate.pay$public.or.private, own.retention.inadequate.pay$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Limited advancement

There is a  relationship between type of ownership and whether an agency indicated that limited advancement opportunities was an obstacle to staff retainment.

Private or non-profit agencies responded that limited advancement was a barrier to retainment at a higher rate than expected - 31 responses (42%) vs. 13 expected (23%). The opposite is true for publicly owned agencies - 100 responded (89%) that limited advancement wasn't a barrier to retainment compared to the 86 expected (77%).

<br>

```{r ownership and retention - limited advancement opportunities}
own.retention.limited.advancement <- own.retention.obstacles %>%
  filter(obstacles == "limited.advancement")

CrossTable(own.retention.limited.advancement$public.or.private, own.retention.limited.advancement$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Lack of recognition

There is no relationship between type of ownership and whether an agency indicated lack of recognition was an obstacle to retainment.

<br>

```{r ownership and retention - lack of recognition}
own.retention.recognition <- own.retention.obstacles %>%
  filter(obstacles == "lack.recognition")

CrossTable(own.retention.recognition$public.or.private, own.retention.recognition$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Community is too isolated

There is no relationship between type of ownership and whether an agency indicated that the community is too isolated and acts as an obstacle to staff retainment.

<br>

```{r ownership and retention - isolated community}
own.retention.isolated.community <- own.retention.obstacles %>%
  filter(obstacles == "community.too.isolated")

CrossTable(own.retention.isolated.community$public.or.private, own.retention.isolated.community$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Employer won't allow staff to leave work

There is a relationship between type of ownership and whether an agency indicated that employers not allowing staff to respond to calls during work hours was an obstacle to staff retainment.

Private or non-profit agencies responded that employers not allowing staff to leave work wasn't a barrier to retention at a higher rate than expected - 55 responses (74%) vs. 46 expected (65%). The opposite is true for publicly owned agencies. 52 of these agencies responded (46%) that employers not allowing staff to leave work was a barrier compared to the expected number of responses - 43 (37%)

<br>

```{r ownership and retention - employer flexibility}
own.retention.employer.flexibility <- own.retention.obstacles %>%
  filter(obstacles == "employer.flexibility")

CrossTable(own.retention.employer.flexibility$public.or.private, own.retention.employer.flexibility$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Too many runs

There is no relationship between type of ownership and whether an agency indicated that too many runs or transports was an obstacle in retainment.

<br>

```{r ownership and retention - too many runs}
own.retention.too.many.runs <- own.retention.obstacles %>%
  filter(obstacles == "too.many.runs")

CrossTable(own.retention.too.many.runs$public.or.private, own.retention.too.many.runs$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Too few runs

There is no relationship between type of ownership and an agency's response to whether too few runs or transports was an obstacle in staff retainment.

<br>

```{r ownership and retention - too few runs}
own.retention.too.few.runs <- own.retention.obstacles %>%
  filter(obstacles == "too.few.runs")

CrossTable(own.retention.too.few.runs$public.or.private, own.retention.too.few.runs$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Dissastisfaction with EMS owners

There is no relationship between type of ownership and an agency's response to whether dissatisfaction with ownership was an obstacle to retainment.

<br>

```{r ownership and retention - dissatisfaction with owners}
own.retention.dissatisfaction.owners <- own.retention.obstacles %>%
  filter(obstacles == "dissatisfaction.owner")

CrossTable(own.retention.dissatisfaction.owners$public.or.private, own.retention.dissatisfaction.owners$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Dissatisfaction with EMS manager

There is no relationship between type of ownership and an agency's response to whether dissatisfaction was an obstacle to staff retainment.

<br>

```{r ownership and retention - dissatisfaction with manager}
own.retention.dissat.manager <- own.retention.obstacles %>%
  filter(obstacles == "dissatisfaction.manager")

CrossTable(own.retention.dissat.manager$public.or.private, own.retention.dissat.manager$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Dissatisfaction with job duties

There is no relationship between type of ownership and an agency's response to whether dissatisfaction with job duties is an obstacle to staff retainment.

<br>

```{r ownership and retention - dissatisfaction with job duties}
own.retention.dissat.duties <- own.retention.obstacles %>%
  filter(obstacles == "dissatisfaction.duties")

CrossTable(own.retention.dissat.duties$public.or.private, own.retention.dissat.duties$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

# Leadership section analysis{.tabset}

The questions in the leadership section of the survey are provided in tabs below along with the chi-square and p-value for the relationship between region and each question.

The questions where region had a relationship in how an agency responded are summarized below.

**Summary:**


*Leaders have training:* 
67% of private/non-profit agencies responded that their leadership was trained compared to 46% for publicly owned agencies.

There were no other questions in which the type of ownership significantly influenced responses.

<br>

## Leaders have training?

**Has the EMS manager of your agency had formal leadership or management training?**

Respondents were provided two choices which are listed below in order of top choice first;

* Yes (54%)
* No (46%)

There is a relationship between type of ownership and whether an agency indicated their managers received leadership and or management training.

Private/non-profit agencies responded that their leadership has received training at a higher rate than expected - 67% yes vs. 33% no.

Publicly owned agencies responded that their leadership has received training at a higher rate than expected - 46% yes vs. 54% no.

<br>

```{r ownership and leadership training}
own.leadership.training <- master.responses %>%
  select(4, 58) %>%
  rename(leadership.training = 2)

CrossTable(own.leadership.training$public.or.private, own.leadership.training$leadership.training, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

## Leadership turnover

**Has your agency had three or more managers in the past five years?**

Respondents were provided two choices which are listed below in order of top choice first;

* No (86%)
* Yes (13%)

There is no relationship between type of ownership and how an agency responded to management turnover.

<br>

```{r ownership and leadership turnover}
own.leadership.turnover <- master.responses %>%
  select(4, 59) %>%
  rename(leadership.turnover = 2)

CrossTable(own.leadership.turnover$public.or.private, own.leadership.turnover$leadership.turnover, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>

## Challenges to management areas{.tabset .tabset-pills}

**Which areas are most difficult to manage?**

Respondents were provided 13 choices and were told to check all that apply. The choices are listed below in order of top choice first;

* Recruitment (69%)
* Staffing the schedule (51%)
* Retention (43%)
* Continuing education requirements (32%)
* Managing staff (31%)
* Policy development and enforcement (25%)
* Staff certification (20%)
* Budget development (17%)
* Billing and reimbursement (15%)
* Maintaining equipment (11%)
* Medical direction (10%)
* None (9%)
* Licensing requirements for the ambulance (5%)

There were zero challenge areas in which the type of ownership seemed to influence responses.



<br>

```{r prep ownership and leadership challenge areas}
own.leadership.challenges <- master.responses %>%
  select(4, 60:74) %>%
  rename(recruitment = 2,
         retention = 3,
         budget = 4,
         staff = 5,
         staffing.schedule = 6,
         medical.direction = 7,
         education.requirements = 8,
         equipment = 9,
         billing = 10,
         licensing = 11,
         staff.certification = 12,
         technology = 13,
         policy = 14,
         technology.2 = 15,
         none = 16) %>%
  gather(key = "challenges", value = "response", 2:16) %>%
  mutate(challenges = str_replace(challenges, "technology.2", "technology"),
         response = ifelse(response == "-", "No", "Yes"))
```

<br>

### Recruitment

There is no relationship between type of ownership and an agencies response to recruitment being a management challenge.

<br>

```{r ownership and leadership challenge areas - recruitment}
own.leadership.challenges.recruitment <- own.leadership.challenges %>%
  filter(challenges == "recruitment")

CrossTable(own.leadership.challenges.recruitment$public.or.private, own.leadership.challenges.recruitment$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Retention

There is no relationship between type of ownership and an agency's response to retention being a management challenge area.

<br>

```{r ownership and leadership challenge areas - retention}
own.leadership.challenges.retention <- own.leadership.challenges %>%
  filter(challenges == "retention")

CrossTable(own.leadership.challenges.retention$public.or.private, own.leadership.challenges.retention$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Budget

There is no relationship between type of ownership and whether an agency identified budget development as a management challenge area.

<br>

```{r ownership and leadership challenge areas - budget}
own.leadership.challenges.budget <- own.leadership.challenges %>%
  filter(challenges == "budget")

CrossTable(own.leadership.challenges.budget$public.or.private, own.leadership.challenges.budget$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Managing staff

There is no relationship between type of ownership and whether an agency identified managing staff as a management challenge area.

<br>

```{r ownership and leadership challenge areas - manageing staff}
own.leadership.challenges.staff <- own.leadership.challenges %>%
  filter(challenges == "staff")

CrossTable(own.leadership.challenges.staff$public.or.private, own.leadership.challenges.staff$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Staffing schedule

There is no relationship between type of ownership and whether an agency identified staffing schedule as an management challenge area.

<br>

```{r ownership and leadership challenge areas - staffing schedule}
own.leadership.challenges.staffing.schedule <- own.leadership.challenges %>%
  filter(challenges == "staffing.schedule")

CrossTable(own.leadership.challenges.staffing.schedule$public.or.private, own.leadership.challenges.staffing.schedule$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Medical direction

There is no relationship between type of ownership and how an agency responded to medical direction as a management challenge area.

<br>

```{r ownership and leadership challenge areas - medical direction}
own.leadership.challenges.medical.direction <- own.leadership.challenges %>%
  filter(challenges == "medical.direction")

CrossTable(own.leadership.challenges.medical.direction$public.or.private, own.leadership.challenges.medical.direction$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Education requirements

There is no relationship between type of ownership and whether an agency indicated that keeping up with education requirements was a management challenge area.

<br>

```{r ownership and leadership challenge areas - education requirements}
own.leadership.challenges.education.req <- own.leadership.challenges %>%
  filter(challenges == "education.requirements")

CrossTable(own.leadership.challenges.education.req$public.or.private, own.leadership.challenges.education.req$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Equipment

There is no relationship between type of ownership and whether an agency indicated that updating equipment was a management challenge.

<br>

```{r ownership and leadership challenge areas - equipment}
own.leadership.challenges.equipment <- own.leadership.challenges %>%
  filter(challenges == "equipment")

CrossTable(own.leadership.challenges.equipment$public.or.private, own.leadership.challenges.equipment$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Billing

There is no relationship between type of ownership and whether an agency indicated that billing and reimbursement was a management challenge area.

<br>

```{r ownership and leadership challenge areas - billing}
own.leadership.challenges.billing <- own.leadership.challenges %>%
  filter(challenges == "billing")

CrossTable(own.leadership.challenges.billing$public.or.private, own.leadership.challenges.billing$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Licensing

There is no relationship between type of ownership and whether an agency indicated that licensing was a management challenge area.

<br>

```{r ownership and leadership challenge areas - licensing}
own.leadership.challenges.licensing <- own.leadership.challenges %>%
  filter(challenges == "licensing")

CrossTable(own.leadership.challenges.licensing$public.or.private, own.leadership.challenges.licensing$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Staff certficiation

There is no relationship between type of ownership and whether an agency indicated that keeping up on staff certification was a management challenge area.

<br>

```{r ownership and leadership challenge areas - staff certification}
own.leadership.challenges.staff.certification <- own.leadership.challenges %>%
  filter(challenges == "staff.certification")

CrossTable(own.leadership.challenges.staff.certification$public.or.private, own.leadership.challenges.staff.certification$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Technology updates

There is no relationship between type of ownership and whether an agency indicated that maintaining technology was a management challenge area.

<br>

```{r ownership and leadership challenge areas - technology}
own.leadership.challenges.technology <- own.leadership.challenges %>%
  filter(challenges == "technology")

CrossTable(own.leadership.challenges.technology$public.or.private, own.leadership.challenges.technology$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### Policy and enforcement

There is no relationship between type of ownership and whether an agency indicated that policy and enforcement was a management challenge area.

<br>

```{r ownership and leadership challenge areas - policy}
own.leadership.challenges.policy <- own.leadership.challenges %>%
  filter(challenges == "policy")

CrossTable(own.leadership.challenges.policy$public.or.private, own.leadership.challenges.policy$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

### None

There is no relationship between type of ownership and whether an agency indicated that there were no management challenge areas.

<br>

```{r ownership and leadership challenge areas - none}
own.leadership.challenges.none <- own.leadership.challenges %>%
  filter(challenges == "none")

CrossTable(own.leadership.challenges.none$public.or.private, own.leadership.challenges.none$response, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)
```

<br>

## Shift Coverage

**What percent of your shifts are fully covered at least 24 hours in advance?**

Respondents were provided 5 choices which are provided below with the top choice listed first;

* 75% - 99% (43%)
* 100% (41%)
* 25% - 49% (3%)
* 0% - 24% (2%)

<br>

There is no relationship between type of ownership and how an agency responded to the percentage of shifts covered 24 hours in advance.



```{r ownership percent shifts covered}
own.shifts.covered <- master.responses %>%
  select(4,75) %>%
  rename(pct.shifts.covered = 2)

CrossTable(own.shifts.covered$public.or.private, own.shifts.covered$pct.shifts.covered, expected = TRUE, prop.t = FALSE, prop.c = FALSE, prop.chisq = FALSE)

```

<br>
